A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-08 tex/sec5-synthesis.tex
[ tex ] Decidiability proof for bidirectional type synthesis and checking before Op

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
We start with the statement of the decidability of bidirectional synthesis and the trichotomy on raw terms for mode-correct bidirectional type systems in \Cref{subsec:trichotomy}, but defer the definition of mode-correctness until \Cref{sec:mode-correctness} and the decidability proof until \Cref{subsec:bidirectional-synthesis-checking}.
We will then formalise mode-correctness and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.

\begin{theorem}[Decidability of Bidirectional Type Synthesis] \label{thm:bidirectional-type-synthesis}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  it is decidable whether there exists some $A$ such that
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A.
  \]
\end{theorem}

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma \not|-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} \not|-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

@@

Now we proceed with the definition of mode-correctness.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                                  & = \emptyset  \\
    \label{eq:synvar2}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \label{eq:synvar3}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align}
    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) ={} & \top \\
    \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
    \label{eq:MC3} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right).
  \end{align}
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
Finally, for a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.

\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from the requirement that each variable in a context is unique.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts : A$ contains a substitution $\rho$ to concrete types from the local context $\Xi$ of an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$.
      To prove that any two derivations for the same $\tmOpts$ has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on $X = \synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ by induction on the list:
      \begin{itemize}
        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then it holds by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
      \end{itemize}
  \end{itemize}
\end{proof}

Uniqueness above is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system, which can involve selecting synthesis arguments whose type variables include the variables of the conclusion.
For instance, for a derivation of $\Gamma |- t\;u \syn B$, one would apply the hypothesis to the sub-derivation $\Gamma |- t \syn A \bto B$.
On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.

\begin{remark}
  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
  With the concept of signature now at our disposal, we can express mode-correctness as defined formally above.
\end{remark}
\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}

For bidirectional type synthesis, we assume further that the set $\Xi$ of variables has a \emph{decidable equality}, rather than being an arbitrary set, so that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.
This assumption is necessary, as type comparisons are ubiquitous throughout type synthesis. 
To simplify our choice, we could just restrict $\Xi$ to any set within the family of sets $\Fin_n$ of natural numbers less than $n$, as defined by \citet{Dybjer1994}.
Given that this set has a decidable equality and the arity of a type construct is always finite, it would be a viable choice.
For clarity in presentation, however, we will continue to use named variables.

\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A;
      \]
    \item if $\erase{\Gamma} |-_{\Sigma, \erase{\Omega}} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis}}]
  We prove this statement by induction on the derivation of $|-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$ (instead of $\isTerm{t}$ alone).
  The first two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness directly.
  The third case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$ to check.
  The last case \Rule{Op} is the interesting part of this theorem.
  \begin{itemize}
    \item \SynRule{Var}: If $\isTerm{t}$ is a variable $\isTerm{x}$ in the context $\Gamma$, then there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.
      Otherwise, if $\isTerm{x}$ is not in $\Gamma$ but $\Gamma |- \isTerm{x} \syn A$ is derivable, then by inversion $\isTerm{x}$ must be in $\Gamma$, a contradiction.
    \item \SynRule{Anno}: For a raw term of the form $\isTerm{t} \annotate A$ in synthesising mode, it is decidable that $\Gamma |- \isTerm{t} \chk A$ is derivable or not by induction hypothesis.
      \begin{enumerate}
        \item If $\Gamma |- \isTerm{t} \chk A$ is derivable, then $\Gamma |- \isTerm{t \annotate A} \syn A$ is derivable.
        \item If $\Gamma |- t \chk A$ is not derivable but $\Gamma |- \isTerm{t \annotate A} \syn$ is derivable, then by inversion $\Gamma |- t \chk A$ is derivable, thus a contradiction.
      \end{enumerate}

    \item \ChkRule{Sub}: If $t$ is in checking mode because of the subsumption rule, then by induction hypothesis it is decidable whether $\Gamma |- t \syn B$ for some $B$:
      \begin{enumerate}
        \item If $\Gamma |- t \syn B$ is derivable for some $B$, then we have to consider any type $A$ and it is decidable if $A$ is equal to $B$:
          \begin{enumerate}
            \item if $A = B$ then we have arrived a derivation of $\Gamma |- t \chk A$.
            \item if $A \neq B$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn A$.
              However, by \Cref{thm:unique-syn}, synthesised types $A$ and $B$ must be equal, so it is a contradiction.
          \end{enumerate}
        \item if $\Gamma |- t \syn C$ is not derivable for any type $C$ but $\Gamma |- t \chk A$ is derivable, then by inversion $\Gamma |- t \syn B$ for some $B = A$, thus a contradiction.
      \end{enumerate}

%    \item For every mode-correct operation $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$, a context $\Gamma$, and a raw term
%      \[
%        \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)}, 
%      \]
%      with a mode $d$, it is decidable that there exists a substitution $\rho$ from $\Xi$ to $\emptyset$ and a typing derivation of
%      \[
%        \Gamma |-_{\Sigma, \Omega} \isTerm{\tmOp_{o}(\vec{x}_1.t_1,\ldots, \vec{x}_n.t_n)} : \simsub{A}{\rho}^{d}.
%      \]
    \item \Rule{Op}:
      Suppose that $t$ is of them $\tmOpts$ for some mode-correct operation $o$ in $\Omega$.
      No matter $t$ is checking or synthesising, we need to build derivations for its arguments incrementally and maintain a set of unifiers, i.e.\ a partial substitution $\rho\colon \PSub{\Xi}{\emptyset}$ is incrementally defined which is initially defined nowhere if $t$ is synthesising or on variables of the target $A$ if $t$ is checking.

  \begin{claim}\label{lem:args-induction}
    For every set $\Xi$ of type variables, list 
    \[
      \mathit{as} = \left([\Delta_i] A_{i}^{\dir{d_i}}\right)_{i = 1}^{n}
    \]
    list of raw terms $t_i$'s in mode $d_i$, context $\Gamma$, and \emph{partial} substitution $\rho$ from $\Xi$ to $\emptyset$
    \begin{enumerate}
      \item either there is a minimal extension $\ext{\rho}$ of $\rho$ defined on $\synvar(\mathit{as})$ such that all of following judgements are derivable
        \[
          \Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\ext{\rho}} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\ext{\rho}}^{\dir{d_i}}
        \]

      \item or there is no extension $\sigma$ of $\rho$ such that all $\Gamma, \vars{x}_\isTerm{i} : \simsub{\Delta_i}{\sigma} |-_{\Sigma, \Omega} \isTerm{t_i} \colon \simsub{A_i}{\sigma}^{\dir{d_i}}$ has a typing derivation. 
    \end{enumerate}
  \end{claim}
  \end{itemize}
\end{proof}

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\LT{trichotomy first or last?}
We start with the statement of the decidability of bidirectional synthesis and the trichotomy on raw terms for mode-correct bidirectional type systems in \Cref{subsec:trichotomy}, but defer the definition of mode-correctness until \Cref{sec:mode-correctness} and the decidability proof until \Cref{subsec:bidirectional-synthesis-checking}.
We will then formalise mode-correctness and show the uniqueness of synthesized types in \Cref{sec:mode-correctness}.
This uniqueness property means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
This will be used later in the proof of the decidability of bidirectional type synthesis and checking in \Cref{subsec:bidirectional-synthesis-checking}.

@@

Now we proceed with the formal definition of mode-correctness.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a signature can be somewhat intricate.
Instead, we choose to define the condition in a few more steps, based on the list of arguments---more specifically, triples of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, and subsequently on a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{variables} which occur in a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by $\fv(\cdot) = \emptyset$ and $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For a list of \emph{arguments} $\overrightarrow{[\Delta_i]A_i^{d_i}}$, define the set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                                  & = \emptyset  \\
    \label{eq:synvar2}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\chk}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \\
    \label{eq:synvar3}\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}] A_{n}^{\dir{\syn}}) & = 
    \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}) \cup \fv(A_{n}).
  \end{align}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness}
  The \emph{mode-correctness for an argument list} $\Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}$  with respect to a subset $S$ of $\Xi$ is defined inductively as
  \begin{align}
    \label{eq:MC1} \MC_{\mathit{as}}( \cdot ) & = \top \\
    \label{eq:MC2} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\syn}\right) 
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right) \right)\\
    \label{eq:MC3} \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}, [\Delta_{n}]A_{n}^{\chk}\right)
    & = \MC_{\mathit{as}}\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)
    \land \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}}\right)\right)
  \end{align}
  where \eqref{eq:MC1} means an empty list is always mode-correct.
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}
  An operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ is \emph{mode-correct} provided that
  \begin{enumerate}
    \item $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$;
    \item $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\overrightarrow{[\Delta_i]A_i^{\dir{d_i}}})$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.
\LT{perhaps we shall prove that mode-correctness is decidable}

\begin{remark}
  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be articulated for individual systems even in the absence of a general definition of bidirectional type systems.
\end{remark}

Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction.
However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall here that a derivation of\/ $\Gamma |- \tmOpts \syn A$ for a synthesis operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\syn}$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types \Cref{fig:extrinsic-typing}.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ agree on variables in $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}}\right)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true by~\eqref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for~\eqref{eq:synvar2} by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness~\eqref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar\left([\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_i]A_{i}^{\dir{d_i}}\right)$, as required for the case \eqref{eq:synvar3} in the definition of $\synvar$.
      \end{enumerate}
  \end{itemize}
\end{proof}

%Uniqueness of the synthesised types is a prevalent property in bidirectional type systems, although the specific proofs can vary depending on the constructs in the system.
%For instance, for derivations of $\Gamma |- t\;u \syn B_i$ for $i = 1, 2$ in simply typed $\lambda$-calculus, the hypothesis is applied to their sub-derivations $\Gamma |- t \syn A_i \bto B_i$ to conclude that $A_1 \bto B_1 = A_2 \bto B_2$ and thus $B_1 = B_2$.
%On the other hand, our proof is based on mode-correctness and need not consider specific sub-derivations.

\subsection{Decidability of Bidirectional Type Synthesis and Checking}\label{subsec:bidirectional-synthesis-checking}
We are ready for the last technical result of this paper---the decidability of bidirectional type synthesis and checking for any mode correct bidirectional type system:
\begin{theorem}[Decidability of Bidirectional Type Synthesis and Checking] \label{thm:bidirectional-type-synthesis-checking}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$,
  \begin{enumerate}
    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$, then it is decidable whether there is a type~$A$ such that
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A;
      \]
    \item if $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\chk}$, then it is decidable for any $A$ whether
      \[
        \Gamma |-_{\Sigma, \Omega} \isTerm{t} \chk A.
      \]
  \end{enumerate}
\end{theorem}

The interesting part of the proof is the case for the $\Rule{Op}$ rule, so we shall give the insight first before jumping into details.
Recall that a typing derivation for $\tmOpts$ for an operation $o\colon \Xi \rhd [\Delta_{1}] A_{1}^{\dir{d}}, \ldots, [\Delta_n]A_{n}^{\dir{d_n}} \to A_0^{\dir{d}}$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$ for $\tmOpts$ while synthesising or checking each argument $t_i$, so we have to start with a \emph{partial} substitution $\rho_0$ defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere, as an `accumulating argument'.
Then, as we visit each synthesis argument of type $A_{i+1}$, we may be able to extend the domain of a partial substitution $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and \emph{extending} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$.
By mode-correctness, the accumulating substitution~$\rho_i$ would be defined on enough synthesis variables so that each type synthesis and checking can be performed on $t_i$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
Then, once we go through every argument $t_i$ successfully, we will have a partial substitution $\rho_n$ defined on every variable $x : \Xi$ by mode-correctness (\Cref{def:mode-correctness}), i.e.\ a substitution $\rho_n$ from $\Xi$ to $\Type_{\Sigma}(\emptyset)$.

To make the aforementioned argument holds, it is necessary to compare types and solve a unification problem.
Hence, we assume that the set $\Xi$ of variables has a decidable equality, rather than being an arbitrary set, thereby ensuring that the set $\Type_{\Sigma}(\Xi)$ of types also has a decidable equality.\footnote{%
To simplify our choice, we could simply confine $\Xi$ to any set within the family of sets $\Fin_n$ of naturals less than~$n$.
Given that these sets have a decidable equality and the arity of a type construct is finite, this would be a viable choice.
Indeed, in our formalisation, we adopt $\Fin_n$ as the set of type variables in the definition of $\Type_{\Sigma}$ (see \Cref{sec:formalisation} for details).
However, for the sake of clarity in presentation, we keep using named variables and just assume that $\Xi$ has a decidable equality.}
The unification problem is as follows:
\begin{lemma}
  For any $A$ of\/ $\Type_{\Sigma}(\Xi)$, $B$ of\/ $\Type_{\Sigma}(\emptyset)$, and a partial substitution $\rho \colon \Xi \to \Type_{\Sigma}(\emptyset)$, 
  \begin{enumerate}
    \item either there is a \emph{minimal extension} $\bar{\rho}$ of $\rho$ such that $\simsub{A}{\bar{\rho}} = B$, or 
    \item there is no extension $\sigma$ of $\rho$ such that $\simsub{A}{\sigma} = B$
  \end{enumerate}
  where by an extension $\sigma \geq \rho$ we mean that the domain of $\sigma$ contains the domain $\rho$ and $\sigma(x) = \rho(x)$ for every $x$ in the domain of $\rho$; by a minimal extension $\bar{\rho}$ of $\rho$ we mean an extension $\bar{\rho}$ of $\rho$ such that $\sigma \geq \bar{\rho}$ whenever $\sigma$ is an extension of $\rho$.
\end{lemma}
This lemma can be inferred from the correctness of first-order unification as proved by \citet{McBride2003,McBride2003a}, or be directly proved without unification.
We now proceed with the decidability proof.

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |- \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  \begin{itemize}
    \item \SynRule{Var}: As $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}$, there exists a type $A$ such that $\Gamma |- \isTerm{x} \syn A$.

    \item \SynRule{Anno}: For $\erase{\Gamma} |-_{\Sigma, \Omega} (\isTerm{t} \annotate A)^{\syn}$, it is decidable whether $\Gamma |- \isTerm{t} \chk A$ by induction hypothesis.
      \begin{itemize}
        \item If $\Gamma |- \isTerm{t} \chk A$, then $\Gamma |- \isTerm{t \annotate A} \syn A$.
        \item If $\Gamma \not|- t \chk A$ but $\Gamma |- \isTerm{t \annotate A} \syn$, then by inversion $\Gamma |- t \chk A$, leading to a contradiction.
      \end{itemize}
